---
title: "1. Concentration-QTc analysis with linear mixed-effects model"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "90%"
)
```

This vignette shows how to perform a concentration-QTc (C-QT) analysis with
`BayesERtools`, using a linear mixed-effects model fitted with
[`rstanarm`](https://mc-stan.org/rstanarm/) (precompiled Stan models; no
compilation required).

C-QT analysis of data from early-phase clinical studies (e.g. single or
multiple ascending dose studies) can be used to exclude a clinically
concerning QTc prolongation, in place of a dedicated thorough QT study
(ICH E14 Q&A). The analysis typically uses the pre-specified linear
mixed-effects model described in the scientific white paper by
[Garnett et al. (2018)](https://doi.org/10.1007/s10928-017-9558-5):

$$
\Delta QTc_{ij} = (\theta_0 + \eta_{0,i}) + \theta_1 TRT_{i} +
(\theta_2 + \eta_{2,i}) C_{ij} + \theta_{3,j} TIME_{j} +
\theta_4 (QTc_{i,baseline} - \overline{QTc_{baseline}}) + \epsilon_{ij}
$$

- $\Delta QTc_{ij}$: change from baseline in QTc for subject $i$ at time $j$
- $\theta_0$, $\eta_{0,i}$: intercept and its between-subject random effect
- $\theta_1$: treatment-specific intercept (active vs placebo)
- $\theta_2$, $\eta_{2,i}$: slope of concentration and its between-subject
  random effect
- $\theta_{3,j}$: time-specific intercept (nominal time as a categorical
  variable)
- $\theta_4$: effect of the baseline QTc (centered by its mean)

The quantity of interest is the placebo-adjusted change from baseline QTc
($\Delta\Delta QTc$) at the clinically relevant concentrations, which is
$\theta_1 + \theta_2 C$ in this model. A QTc effect exceeding 10 ms can be
excluded if the upper bound of the two-sided 90% confidence interval of
$\Delta\Delta QTc$ is below 10 ms at the highest clinically relevant
exposure. With the Bayesian approach, we can in addition directly quantify
the posterior probability that $\Delta\Delta QTc$ exceeds 10 ms.

```{r setup, message = FALSE}
library(BayesERtools)
library(dplyr)
library(ggplot2)
theme_set(theme_bw(base_size = 12))
```

## Data

We use `d_sim_cqt`, a simulated dataset of a hypothetical single ascending
dose study with 6 dose cohorts (25 to 800 mg). Each cohort has 6 subjects on
active drug and 2 on placebo, with time-matched PK and QTcF measurements at
10 time points up to 24 hours post-dose. We assume that 200 mg is the
therapeutic dose and 800 mg is the supratherapeutic dose. See
`?d_sim_cqt` for details.

```{r data}
d_sim_cqt
```

Concentrations are provided in ng/mL (`CONC`) and ug/mL (`CONC_1000`). We
use ug/mL for the analysis, as the prior on the random effects covariance
is not scaled to the units of the exposure; badly scaled exposure (e.g.
ng/mL, ranging up to thousands) can lead to poor sampling when a random
slope is included (`dev_ermod_cqt()` warns in such cases).

Let's first look at the time course of concentration and $\Delta QTcF$.

```{r data-time-course, fig.height = 6}
d_mean_time <-
  d_sim_cqt |>
  mutate(DOSE = factor(DOSE)) |>
  summarize(
    CONC_1000 = mean(CONC_1000),
    DQTCF = mean(DQTCF),
    .by = c(DOSE, NTIME)
  ) |>
  tidyr::pivot_longer(c(CONC_1000, DQTCF))

ggplot(d_mean_time, aes(NTIME, value, color = DOSE)) +
  geom_line() +
  geom_point() +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Nominal time (h)", y = "Mean value", color = "Dose (mg)")
```

## Model development

`dev_ermod_cqt()` builds the model above from the specified variables.
The treatment, time, and baseline terms are included if `var_trt`,
`var_time`, and `var_baseline` are specified, respectively. By default,
correlated random effects on the intercept and slope are included
(`random_effect_type = "both"`).

Mixed-effects models sometimes show warnings about divergent transitions
during sampling. Here we increase `adapt_delta` from the default of 0.95 to
0.99 to avoid them (at the cost of slightly longer run time).

```{r dev-mod}
set.seed(1234)
ermod_cqt <- dev_ermod_cqt(
  data = d_sim_cqt,
  var_resp = "DQTCF",
  var_exposure = "CONC_1000",
  var_random = "ID",
  var_trt = "TRT",
  var_time = "NTIME",
  var_baseline = "QTCFBL",
  adapt_delta = 0.99
)

ermod_cqt
```

The priors used can be checked with `prior_summary()`. Weakly informative
priors of `rstanarm` are used by default, and can be modified with the
`prior`, `prior_intercept`, `prior_aux`, and `prior_covariance` arguments.

```{r prior}
prior_summary(ermod_cqt)
```

## Model assessment

### Hysteresis

The model assumes a direct effect of concentration on QTc. A hysteresis plot
shows the mean placebo-adjusted $\Delta QTcF$ ($\Delta\Delta QTcF$) against
the mean concentration at each time point, connected in the order of time
(the numbers show the nominal time). A systematic counter-clockwise loop
would indicate a delayed effect. With only 6 subjects per dose the means are
noisy, but there is no such loop here: $\Delta\Delta QTcF$ peaks around the
time of peak concentration (1 to 3 hours). Use `var_group` (e.g.
`var_group = "DOSE"`) to show the plot by dose.

```{r hysteresis}
plot_cqt_hysteresis(ermod_cqt)
```

### Goodness-of-fit

The goodness-of-fit plot compares the model prediction of
$\Delta\Delta QTcF$ with the observed $\Delta\Delta QTcF$ (time-matched
placebo-adjusted) in concentration deciles. The observed data are well
described by the linear model, supporting the linearity assumption.

```{r gof}
plot_cqt_gof(ermod_cqt)
```

Residuals of individual predictions (using each subject's posterior random
effects; see [Simulation](#simulation) below) can be checked with
`sim_er()`.

```{r resid, fig.height = 3.5}
ersim_ind <- sim_er(
  ermod_cqt,
  n_draws_sim = 500,
  re_type = "existing_subject", output_type = "median_qi"
)

ersim_ind |>
  mutate(.resid = DQTCF - .epred) |>
  ggplot(aes(CONC_1000, .resid)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  labs(y = "Residual (individual prediction)")
```

## Predicted $\Delta\Delta QTcF$ at clinically relevant concentrations

The geometric mean Cmax for each dose level is calculated from the data.

```{r cmax}
d_cmax <-
  d_sim_cqt |>
  filter(TRT == "Active") |>
  summarize(CMAX = max(CONC_1000), .by = c(ID, DOSE)) |>
  summarize(GM_CMAX = exp(mean(log(CMAX))), .by = DOSE)

d_cmax
```

`sim_cqt_ddqtc()` calculates the posterior median and credible interval
(90% by default) of $\Delta\Delta QTcF$ at the specified concentrations,
along with the posterior probability of $\Delta\Delta QTcF$ exceeding the
threshold (10 ms by default).

```{r ddqtc}
cmax_to_eval <- setNames(d_cmax$GM_CMAX, paste(d_cmax$DOSE, "mg"))

d_ddqtc <- sim_cqt_ddqtc(ermod_cqt, cmax_to_eval)

d_ddqtc |>
  gt::gt() |>
  gt::fmt_number(columns = c(CONC_1000, .ddqtc, .lower, .upper), n_sigfig = 3) |>
  gt::fmt_percent(columns = .prob_exceed, decimals = 1) |>
  gt::cols_hide(.width) |>
  gt::cols_label(
    .label = "Dose",
    CONC_1000 = "GM Cmax (ug/mL)",
    .ddqtc = "ΔΔQTcF (ms)",
    .lower = "90% CrI lower",
    .upper = "90% CrI upper",
    .prob_exceed = "Pr(ΔΔQTcF > 10 ms)"
  )
```

At the therapeutic dose (200 mg), the upper bound of the 90% credible
interval is well below 10 ms, and the posterior probability of exceeding
10 ms is negligible. At the supratherapeutic dose (800 mg), a QTc effect
above 10 ms can not be excluded.

The predictions can be shown on the goodness-of-fit plot with
`exposure_to_mark`.

```{r gof-mark}
plot_cqt_gof(ermod_cqt, exposure_to_mark = cmax_to_eval[c("200 mg", "800 mg")])
```

The posterior draws are also available with `output_type = "draws"`, e.g. to
visualize the posterior distribution.

```{r ddqtc-draws, fig.height = 3.5}
sim_cqt_ddqtc(
  ermod_cqt, cmax_to_eval[c("200 mg", "800 mg")],
  output_type = "draws"
) |>
  ggplot(aes(.ddqtc, fill = .label)) +
  ggdist::stat_slab(alpha = 0.6) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  labs(x = "ΔΔQTcF (ms)", y = "Posterior density", fill = NULL)
```

## Simulation {#simulation}

Simulation functions such as `sim_er()`, `sim_er_new_exp()`, and
`sim_er_curve()` support three types of random effects with the `re_type`
argument for mixed-effects models:

- `"population"` (default): random effects set to zero, i.e. prediction for
  a typical subject
- `"new_subject"`: random effects sampled from the estimated between-subject
  distribution, for each subject (`ID`) in the simulation dataset
- `"existing_subject"`: each subject's own posterior random effects
  (individual prediction)

Note that the model includes the treatment, time, and baseline terms
(with derived columns `TRT_active`, `NTIME_fct`, and `QTCFBL_cent`; see
`?dev_ermod_cqt`), which need to be included in the simulation dataset.
Here we simulate $\Delta QTcF$ of 100 new subjects on active treatment at
2 h post-dose with the mean baseline QTcF.

```{r sim-new-subject}
data_cov_new <- tibble(
  ID = 1:100,
  TRT_active = 1,
  NTIME_fct = factor(2, levels = levels(extract_data(ermod_cqt)$NTIME_fct)),
  QTCFBL_cent = 0
)

ersim_new <- sim_er_curve(
  ermod_cqt,
  exposure_range = c(0, 5),
  num_exposures = 11,
  data_cov = data_cov_new,
  n_draws_sim = 200,
  re_type = "new_subject"
)

# Distribution of the expected dQTcF across new subjects and posterior draws
ersim_new |>
  summarize(
    .epred_median = median(.epred),
    .epred_lower = quantile(.epred, 0.05),
    .epred_upper = quantile(.epred, 0.95),
    .by = CONC_1000
  ) |>
  ggplot(aes(CONC_1000, .epred_median)) +
  geom_ribbon(aes(ymin = .epred_lower, ymax = .epred_upper), alpha = 0.3) +
  geom_line() +
  labs(
    y = "ΔQTcF (ms)",
    caption = "Line/shade: median/90% interval across new subjects and posterior draws"
  )
```

## Model comparison

The random slope on concentration can be difficult to estimate with a
limited number of subjects. A model with only a random intercept can be
compared with `loo()`. Similarly, a model with the metabolite
concentration can be evaluated to check which exposure metric better
describes the QTc data.

```{r loo, warning = FALSE}
set.seed(1234)
ermod_cqt_int <- dev_ermod_cqt(
  data = d_sim_cqt,
  var_resp = "DQTCF",
  var_exposure = "CONC_1000",
  var_random = "ID",
  var_trt = "TRT",
  var_time = "NTIME",
  var_baseline = "QTCFBL",
  random_effect_type = "intercept",
  adapt_delta = 0.99,
  verbosity_level = 0
)

ermod_cqt_metab <- dev_ermod_cqt(
  data = d_sim_cqt,
  var_resp = "DQTCF",
  var_exposure = "MCONC_1000",
  var_random = "ID",
  var_trt = "TRT",
  var_time = "NTIME",
  var_baseline = "QTCFBL",
  adapt_delta = 0.99,
  verbosity_level = 0
)

loo::loo_compare(list(
  parent_int_slope = loo(ermod_cqt),
  parent_int = loo(ermod_cqt_int),
  metabolite = loo(ermod_cqt_metab)
))
```

The models with the parent concentration have substantially higher
expected log predictive density (ELPD) than the one with the metabolite
concentration. The two models with the parent concentration perform
similarly, and the conclusion on $\Delta\Delta QTcF$ is robust to the
choice of the random effects structure:

```{r ddqtc-int}
sim_cqt_ddqtc(ermod_cqt_int, cmax_to_eval[c("200 mg", "800 mg")])
```

## Reference

Garnett C, Bonate PL, Dang Q, et al. Scientific white paper on
concentration-QTc modeling. J Pharmacokinet Pharmacodyn. 2018;45(3):383-397.
<https://doi.org/10.1007/s10928-017-9558-5>
